\mbox{Specifying Imperative Higher Order Programs}

### 2009-10-01 papers/esop2010/main.tex
biblio, and a lot of text (7 pages)

\newcommand{\wpre}{ {\bf wp}}
\newcommand{\sep}{ ~|~ }
\newcommand{\letml}{{\bf let}}
\newcommand{\inml}{{\bf in}}
\newcommand{\ifml}{{\bf if}}
\newcommand{\thenml}{{\bf then}}
\newcommand{\elseml}{{\bf else}}
\newcommand{\refml}{{\bf ref}}
\newcommand{\propml}{{\bf prop}}
\newcommand{\boolml}{{\bf bool}}
\newcommand{\recml}{{\bf rec}}
\newcommand{\efft}[1]{ \langle #1 \rangle }
\newcommand{\lift}[1]{ \langle #1 \rangle }
\newcommand{\alist}[1]{ \overline{#1} }
\newcommand{\Trueml}{{\bf True}}
\newcommand{\Falseml}{{\bf False}}
\newcommand{\trueml}{{\bf true}}
\newcommand{\falseml}{{\bf false}}
\newcommand{\unitml}{{\bf unit}}
\newcommand{\coq}{Coq}

@@

\author{Johannes Kanig \and Jean-Christophe Fill\^atre}
\institute{
Laboratoire de Recherche en Informatique, Univ Paris-Sud, CNRS, Orsay F-91405\\
INRIA Futurs, ProVal, Parc Orsay Universit\'e, F-91893
}
%TODO affiliation

@@

\section{Introduction.}
%TODO Hoare logic = sharing between logic and prog language + alias free

@@

\section{A programming language for specification.}

@@

Our language is basically λ-calculus with \letml-bindings. For technical
reasons, the presentation is different from its usual form.
Fig.\ref{fig:syntax} describes the whole language.

Let us first look at the {\em types}. Type variables are of course types, as
well as tuples ($τ\times τ$) and pure functions ($τ -> τ$). There is a second
function type $τ->^ε τ$ for impure functions, annotated with an {\em effect}
$ε$. Effects are simply lists of {\em regions} $ρ$ and {\em effect variables}
$e$. The type expression $τ~\refml_ρ$ describes the type of references in
region $ρ$ containing objects of type $τ$. The type $\efft{ε}$ describes the
portion of the store represented by the effect expression $ε$.

The syntax of terms contains some differences to standard λ-calculus. The
construct $λx:τ.t$ serves to construct a pure function, while the construct
$\recml~f~(x:τ).\{p\}t\{q\}$ can be used to build impure, possibly recursive
functions. The terms $p$ and $q$ are the {\em precondition} and the {\em
postcondition} of the so defined anonymous function. This is of course
reminiscent of the syntax of Hoare Logic.

Type, region and effect variables must be explicitly generalized and
instantiated. For this purpose, we have defined the metavariable $χ$,
representing variables, and $κ$, representing the corresponding instantiation
objects types, regions and effects\footnote{We do not distinguish between
region variables and region constants; a region constant is simply a region
variable whose scope is the entire program.}. Now, it can be seen that
generalization takes place at \letml-bindings and instantiation at variable
occurrences.
%TODO describe the overline

%TODO : get rid of instantiations ??
\subsection{Typing}

\begin{figure}[htpb]
\begin{mathpar}
  { \inferrule*[Left=Var] {Γ(x) = ∀\alist{χ}.τ} {Γ|-_v x~\alist{κ} :
  τ[\alist{χ}|->\alist{κ}]} } \and
  { \inferrule*[Left=PureFun] {Γ,x : τ' |-_v t : τ} {Γ|-_v λx : τ'.t : τ' ->
  τ} } \and
  { \inferrule*[Left=Rec] 
  { Γ' = Γ,~f : τ' ->^ε τ,~ x : τ' \\ Γ' |- t : τ,ε \\ Γ', cur : \efft{ε} |-_l p : \propml \\ 
  Γ', old, cur : \efft{ε} |-_l q : τ -> \propml } 
      {Γ|-_v \recml~f~(x : τ').\{p\}t\{q\} : τ' ->^ε τ} 
  } \and
  { \inferrule*[Left=PureApp]
    {Γ |-_v t1 : τ' -> τ \\ Γ|-_v t2 : τ'}
    {Γ |-_v  t1 ~ t2 : τ}
  } \and
  { \inferrule*[Left=App]
    {Γ |-_v t1 : τ' ->^ε τ \\ Γ|-_v t2 : τ'}
    {Γ |- t1 ~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=ite]
    { Γ |- _v t1 : \boolml \\ Γ |- t2 : τ, ε1 \\ Γ |- t3 : τ, ε2 }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ, ε1  ε2 }
  } \and
  { \inferrule*[Left=Letv]
    { Γ, \alist{χ} |-_v t1 : τ' \\ Γ, x : ∀\alist{χ}.τ' |- t2 : τ,ε }
    { Γ |- \letml~ x~[\alist{χ}] = t1 ~ \inml~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=Let]
    { Γ |- t1 : τ', ε1 \\ Γ, x : τ' |- t2 : τ, ε2 }
    { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε1  ε2 }
  } \and
  { \inferrule*[Left=Value]
  { Γ |-_v t : τ  }
  { Γ |- t : τ,\emptyset  }
  } \and
  { \inferrule*[Left=Logic]
    { Γ |-_v t : τ  }
    { Γ |-_l t : \ceil{τ} }
  }
\end{mathpar}
  \caption{The Typing relation}
  \label{fig:typing}
\end{figure}

@@

Again in order to simplify future definitions, the definition typing relation
differs somewhat from the standard way (see Fig.~\ref{fig:typing}). First of
all, we present the language in {\em A-normal form} (ANF), where all values
obtained by intermediate computations are named. However, our definition of
values is not syntactical: our ``semantic'' values are objects which are
either syntactic values (abstractions and variables) or are obtained by {\em
pure application}, the application which involves the pure arrow $τ->τ$.
Furthermore, the system is a type and effect system. There are in fact two
typing relations, the first one is of the form $Γ|-_v  t : τ$. In this case,
$t$ is a semantic value of type $τ$ and cannot have any effect. It can be seen
that the set of terms described by $|-_v$ is the precisely the one obtained by
combining syntactic values and pure application in a well-typed way.

@@

unusual : 
missing : 
* value injection
* subeffecting
* logic typing

@@

\subsection{Sharing between logic and programs}

To obtain sharing between
\subsection{References}

@@

\begin{figure}[htpb]
  \begin{mathpar}
    {\inferrule*[Left=Var]
      {Γ(x) = ∀\alist{χ}.τ \\ σ = [\alist{χ}|->\alist{κ}] \\ σ \sim τ }
      {Γ |- x : τσ }
    } \\
    { \inferrule*[Left=LetRef, Right={ρ \textup{fresh}, $ρ\notin τ$}]
      {Γ |-_v t1 : τ' \\ Γ, x : τ'~\refml_ρ |- t2 : τ }
      {Γ |- \letml~x = ref~ t1 ~\inml~ t2 : τ}
    }
  \end{mathpar}
  \caption{A modified and a new rule for aliasing}
  \label{fig:aliasing}
\end{figure}

\section{A Simple Weakest Preconditions Calculus.}

@@

\section{Extensions.}

@@

\section{Implementation.}

\section{Examples.}

\section{Related Work.}

\section{Conclusion.}

>>>>

\newcommand{\wpre}{ {\bf wp}}
\newcommand{\sep}{ ~|~ }
\newcommand{\letml}{{\bf let}}
\newcommand{\inml}{{\bf in}}
\newcommand{\ifml}{{\bf if}}
\newcommand{\thenml}{{\bf then}}
\newcommand{\elseml}{{\bf else}}
\newcommand{\refml}{{\bf ref}}
\newcommand{\propml}{{\bf prop}}
\newcommand{\boolml}{{\bf bool}}
\newcommand{\recml}{{\bf rec}}
\newcommand{\efft}[1]{ \langle #1 \rangle }
\newcommand{\lift}[1]{ \langle #1 \rangle }
\newcommand{\alist}[1]{ \overline{#1} }
\newcommand{\Trueml}{{\bf True}}
\newcommand{\Falseml}{{\bf False}}
\newcommand{\trueml}{{\bf true}}
\newcommand{\falseml}{{\bf false}}
\newcommand{\unitml}{{\bf unit}}
\newcommand{\coq}{Coq}
\newcommand{\who}{Who}
\newcommand{\correct}[1]{ c(#1) }

@@

\author{Johannes Kanig \and Jean-Christophe Fill\^atre}
\institute{
Laboratoire de Recherche en Informatique, Univ Paris-Sud, CNRS, Orsay F-91405\\
INRIA Futurs, ProVal, Parc Orsay Universit\'e, F-91893
}
%TODO affiliation
%TODO find notions for:
%  * ``semantic values'' v
%  * ``store types'' <ε>
%

@@

\section{Introduction}
%TODO Hoare logic = sharing between logic and prog language + alias free

@@

\section{A programming language for specification}

@@

Our programming language (Fig.~\ref{fig:syntax}) is basically the λ-calculus
with \letml-bindings, where there are two different λ-bindings. The first one
($λx:τ.t$) describes a pure, non-recursive binding, and the second one
($\recml~f~(x:τ).\{p\}t\{q\}$) describes recursive, potentially impure
functions; $p$ and $q$ in this term are {\em specifications}, which are
of the same syntax as programs, but are differently typed (typing will be
discussed in the next section). The term $p$ corresponds to the {\em
precondition}, $q$ to the {\em postcondition} of the function. The analogy to
Hoare triples is of course obvious. As usual, there are also variables,
\letml-bindings and an \ifml-\thenml-\elseml-construct. Variables on the type
level can be generalized at \letml-boundaries and have to be instantiated when
the corresponding program variable is used.

Before we describe the syntax of types, let us first explain {\em regions} and
{\em effects}. Regions are simply type level names for memory locations. An
effect is a set of memory locations and {\em effect variables}. An effect
variable stands for an unknown effect. Effects are used to specify
of memory locations an expression can write. They are also used in function
types to describe the memory locations which may be written if the function is
called.

Type, region and effect variables must be explicitly generalized and
instantiated. For this purpose, we have defined the metavariable $χ$,
representing variables, and $κ$, representing the corresponding instantiation
objects types, regions and effects\footnote{We do not distinguish between
region variables and region constants; a region constant is simply a region
variable whose scope is the entire program.}. Now, it can be seen that
generalization takes place at \letml-bindings and instantiation at variable
occurrences. We use a horizontal bar over terms (like in ${\overline{χ}}$) to
denote lists.

\subsection{Typing}

\begin{figure}[htbp]
  \begin{mathpar}
  { \inferrule*[Left=Var] {Γ(x) = ∀\alist{χ}.τ} {Γ|-_v x~\alist{κ} :
  τ[\alist{χ}|->\alist{κ}]} } \and
  { \inferrule*[Left=PureFun] {Γ,x : τ' |-_v t : τ} {Γ|-_v λx : τ'.t : τ' ->
  τ} } \and
  { \inferrule*[Left=Rec] 
  { Γ' = Γ, x : τ' \\ Γ'' = Γ',~f : τ' ->^ε τ \\ Γ'' |- t : τ,ε \\ Γ', cur : \efft{ε} |-_l p : \propml \\ 
  Γ', old, cur : \efft{ε} |-_l q : τ -> \propml } 
      {Γ|-_v \recml~f~(x : τ').\{p\}t\{q\} : τ' ->^ε τ} 
  } \and
  { \inferrule*[Left=PureApp]
    {Γ |-_v t1 : τ' -> τ \\ Γ|-_v t2 : τ'}
    {Γ |-_v  t1 ~ t2 : τ}
  }
  \end{mathpar}
  \caption{Value typing}
  \label{fig:valuetyping}
\end{figure}

\begin{figure}[htpb]
\begin{mathpar}
  { \inferrule*[Left=Value]
  { Γ |-_v t : τ  }
  { Γ |- t : τ,\emptyset  }
  } \and
  { \inferrule*[Left=App]
    {Γ |-_v t1 : τ' ->^ε τ \\ Γ|-_v t2 : τ'}
    {Γ |- t1 ~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=ite]
    { Γ |- _v t1 : \boolml \\ Γ |- t2 : τ, ε1 \\ Γ |- t3 : τ, ε2 }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ, ε1  ε2 }
  } \and
  { \inferrule*[Left=Letv]
    { Γ, \alist{χ} |-_v t1 : τ' \\ Γ, x : ∀\alist{χ}.τ' |- t2 : τ,ε }
    { Γ |- \letml~ x~[\alist{χ}] = t1 ~ \inml~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=Let]
    { Γ |- t1 : τ', ε1 \\ Γ, x : τ' |- t2 : τ, ε2 }
    { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε1  ε2 }
  }
\end{mathpar}
  \caption{Program typing}
  \label{fig:progtyping}
\end{figure}

\begin{figure}[htbp]
  \begin{mathpar}
  { \inferrule*[Left=Logic]
    { Γ |-_v t : τ  }
    { Γ |-_l t : \ceil{τ} }
  } \and
  { \inferrule*[Left=LetLogic]
    { Γ, \alist{χ} |-_l : t' : τ' \\ Γ, x: ∀\alist{χ}.τ' |- t : τ  }
    { Γ |-_l \letml~x~[\alist{χ}] = t'~\inml~t : τ }
  }
  \end{mathpar}
  \caption{Logic typing}
  \label{fig:logictyping}
\end{figure}

@@

\begin{figure}[htbp]
  \begin{eqnarray*}
    \ceil{x~[\alist{κ}]} &=& x~[\alist{κ}]\\
    \ceil{λx:τ.t} &=& λx:τ.\ceil{t}\\
    \ceil{\recml~f~(x:τ).\{p\} (t : ε ) \{q\}} &=& 
    (λx:\ceil{τ}.λcur:\efft{ε}.p,λx:\ceil{τ}.λold,cur:\efft{ε}.q)\\
    \ceil{t_{τ' -> τ}~t'} &=& \ceil{t}~\ceil{t'}
  \end{eqnarray*}
  \caption{Lifting values}
  \label{fig:valuelift}
\end{figure}

We have already seen that programs and specification share in fact the same
syntax. Though they are differently typed, their typing relations still share
a lot. For this reason, and to underline the fact that in Hoare logic, values
are shared between programs and the logic, we present three different typing
relations. One for programs, one for specifications, and the common part is
factored out in a third typing relation for values.

Again in order to simplify future definitions, the definition typing relation
differs somewhat from the standard way (see Fig.~\ref{fig:progtyping}). First of
all, we present the language in {\em A-normal form} (ANF), where all values
obtained by intermediate computations are named. However, our definition of
values is not syntactical: our ``semantic'' values are objects which are
either syntactic values (abstractions and variables) or are obtained by {\em
pure application}, the application which involves the pure arrow $τ->τ$.
Furthermore, the system is a type and effect system. There are in fact two
typing relations, the first one is of the form $Γ|-_v  t : τ$. In this case,
$t$ is a semantic value of type $τ$ and cannot have any effect. It can be seen
that the set of terms described by $|-_v$ is precisely the one obtained by
combining syntactic values and pure application in a well-typed way.

@@

missing : 
* subeffecting

@@

\subsection{predefined constants}

arithmetic operations, operations on pairs, logic operations on prop, pre/post

reference operations
!!

operations on maps : restrict, combine
\subsection{Sharing between logic and programs}

use logic application in programs : no problem;
use lambdas with effects in logic : specification

\subsection{References}

@@

\begin{figure}[htpb]
  \begin{mathpar}
    {\inferrule*[Left=Var]
      {Γ(x) = ∀\alist{χ}.τ \\ σ = [\alist{χ}|->\alist{κ}] \\ σ \sim τ }
      {Γ |- x : τσ }
    } \\
    { \inferrule*[Left=LetRef, Right={ρ \textup{fresh}, $ρ\notin τ$}]
      {Γ |-_v t1 : τ' \\ Γ, x : τ'~\refml_ρ |- t2 : τ, ε }
      {Γ |- \letml~x = ref~ t1 ~\inml~ t2 : τ, ε \setminus ρ}
    }
  \end{mathpar}
  \caption{A modified and a new rule for aliasing}
  \label{fig:aliasing}
\end{figure}

\subsection{Some properties}

$Γ|-_v v : τ$ implies $Γ |-_l \ceil{v} : \ceil{τ}$

\section{A Simple Weakest Preconditions Calculus}

@@

\begin{figure}[htbp]
  \begin{eqnarray*}
    \wpre_s(v, q) &=& q~s~\ceil{v}\\
    \wpre_s( t_{τ'->^ε τ} t' ,q) &=& \correct{t} /\ \correct(t') /\
    fst~\ceil{t}~\ceil{t'}~s /\ \\ 
    & & ∀s':\efft{ε}, x:τ.~  snd~\ceil{t}~\ceil{t'}~s~s'~x => q~s'~x\\
    \wpre_s(\letml~x~[\alist{χ}] = v~\inml~t, q) &=&
      c(v) /\ \letml~x~[\alist{χ}]=\ceil{v}~\inml~ \wpre(t,q)\\
    \wpre_s(\letml~x = ( t1 : ε1 ) ~\inml~ ( t2 : ε2 ) , q) &=&
    \wpre_s( t1 ,λs':\efft{ ε1  ε2 }.\wpre_{s'}(t2, q ))\\
    \wpre_s(\ifml~ t1 ~\thenml~ t2 ~\elseml~ t3 , q) &=&
    (\ceil{ t1 } = \trueml => \wpre_s( t2 , q)) /\ \\
    & &(\ceil{ t1 } = \falseml => \wpre_s( t3 , q)) 
  \end{eqnarray*}
  \caption{The weakest precondition calculus}
  \label{fig:wp}
\end{figure}

\section{Extensions}

@@

\section{Implementation}

%what has been implemented

A prototype implementation by the name ``\who'' exists and is freely available. It implements the
described language with type, region and effect polymorphism, higher-order
logics and store types in the logic. It takes as an input a fully annotated
program and outputs proof obligations which have to be proved to guarantee the
correctness of the program with respect to its specification. Currently, only
output to the interactive proof assistant Coq has been implemented. Our logic
is a subset of Coq's, so the translation is particularly simple.

%details
% lines of code : less than 3000 lines of ocaml
% limited form of type and effect inference
%

\section{Examples}

\begin{center}
    \begin{tabular}{ | r | r | r | r |}
    \hline
    Program & Code & Spec & Proofs \\ \hline
    Array &  &  &  \\
    Hashtbl & & &  \\ \hline
    Memo/Ymemo & & &  \\ \hline
    Koda-Ruskey~\cite{KanigFilliatre09wml} & & &  \\ \hline
    \end{tabular}
\end{center}

\subsection{memo}
\subsection{ymemo}

\section{Related Work}

\section{Conclusion}

\paragraph{Future Work.}

We are currently investigating the possibility to obtain first-order proof
obligations, to be able to use SMT solvers~\cite{RanTin-SMTLIB}, which would
avoid manual proofs in many cases. An obvious solution would be to reuse
existing encodings of higher-order logic in first-order logic, such as the one
by Pottier and Gauthier~\cite{pottier-gauthier-hosc}. This technique has been
successfully employed by the Pangolin system~\cite{regis-gianas-pottier-08}.
Another option comes from our observation that in many cases, the proof
obligations for higher-order functions can actually be expressed in
first-order logic, when one replaces top-level {\em quantifications} of
functional arguments and invariants by top-level {\em declarations}. Doing
this translation manually, we have obtained first-order proof
obligations for the {\tt Array.iter} function which are all discharged by
SMT solvers. We don't know if such a transformation can always be carried out.

what about correspondence between $τ ->^\emptyset τ$ and $τ -> τ$?

\bibliographystyle{plain}
\bibliography{biblio}{}
